---
title: "OA Dashboard"
format: 
  dashboard:
    nav-buttons: [github]
    github: https://github.com/tesaunders/oa-nz
    output-file: index.html
logo: images/oa.png
theme:
  - darkly
fig-width: 12
---

```{r}
#| label: load-packages
#| message: false

library(dplyr)
library(purrr)
library(ggplot2)
library(ggthemes)
library(gt)
library(tidyr)
library(scales)
library(forcats)
```

```{r}
#| label: define-values-themes

prev_year <- as.numeric(format(Sys.Date(), "%Y"))-1

plot_theme <- function(base_size=20, base_family="sans") {
  (theme_foundation(base_size=base_size, base_family=base_family)
   + theme(plot.title = element_text(face = "bold", colour = 'white',
                                    hjust = 0.5, margin = margin(0,0,20,0)),
           text = element_text(),
           panel.background = element_rect(colour = NA, fill = 'grey20'),
           plot.background = element_rect(colour = NA, fill = '#2d2d2d'),
           panel.border = element_rect(colour = NA),
           axis.title = element_text(face = "bold",size = rel(1), colour = 'white'),
           axis.title.y = element_text(angle=90,vjust =2),
           axis.title.x = element_text(vjust = -0.2),
           axis.text = element_text(colour = 'grey70'),
           axis.line.x = element_line(colour="grey70"),
           axis.line.y = element_line(colour="grey70"),
           axis.ticks = element_line(colour="grey70"),
           panel.grid.major = element_line(colour="#262626"),
           panel.grid.minor = element_blank(),
           legend.background = element_rect(fill ='#262626'),
           legend.text = element_text(color = 'white'),
           legend.key = element_rect(colour = NA, fill = '#262626'),
           legend.position = "right",
           legend.direction = "vertical",
           legend.box = "vertical",
           legend.key.size= unit(0.5, "cm"),
           legend.title = element_blank(),
           plot.margin=unit(c(10,5,5,5),"mm"),
           strip.background=element_rect(colour="#2D3A4C",fill="#2D3A4C"),
           strip.text = element_text(face="bold", colour = 'white')))
}

plot_colours <- c("bronze" = "#A65628",
                  "hybrid" = "#377EB8",
                  "gold" = "#FFFF33",
                 "diamond" = "lightgrey",
                 "green" = "#4DAF4A",
                 "closed" = "#E41A1C")
```

```{r}
#| label: load-data

pub_data <- read.csv("data/publications.csv")
cit_nz <- read.csv("data/citations-nz.csv")
cit_au <- read.csv("data/citations-au.csv")

oa_levels <- c("closed", "hybrid", "gold", "green", "diamond", "bronze")

pub_data$key <- factor(pub_data$key, levels = oa_levels)

cit_nz$oa_status <- factor(cit_nz$oa_status, levels = oa_levels)
cit_au$oa_status <- factor(cit_au$oa_status, levels = oa_levels)
```

#  {.sidebar}

This dashboard displays data relating to open access publication output from all 8 Aotearoa New Zealand universities and the Australian 'Group of 8' universities.

------------------------------------------------------------------------

Data is extracted from [OpenAlex](https://openalex.org/about) via the [API](https://docs.openalex.org/) to examine publication trends since `r min(pub_data$year)`, and citation trends since `r min(cit_nz$publication_year)`. Currently, data is only included up to the most recent full year (`r prev_year`). Data and code are available on [GitHub](https://github.com/tesaunders/oa-nz/).

------------------------------------------------------------------------

Created and maintained by [Tom Saunders](https://tomsaunders.me).

------------------------------------------------------------------------

::: {.callout-note collapse="true"}
## Notes on data

Data includes all journal articles, books, and book chapters with an author affiliated with at least one of the institutions named above. Works about the publication venue ("paratext") have been excluded (e.g. covers, table of contents, editorial board listing, issue information), as have retracted works.

:::

```{r}
#| label: summarise-data

total <- 
  pub_data |> 
  mutate(
    open = case_when(key != "closed" ~ "open", .default = key),
  ) |> 
  filter(open == "open") |> 
  group_by(year, country, institution) |> 
  summarise(
    pc = sum(pc),
  ) |> 
  group_by(year, country) |> 
  summarise(
    pc = mean(pc),
  )
```

# Publications (NZ)

## Row {height=20%}

```{r}
#| content: valuebox
#| title: "Current OA rate"

total_nz <- 
  total |> 
  filter(country == "nz", year == prev_year) |> 
  pull(pc)

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_nz), "%")
)
```

```{r}
#| content: valuebox
#| title: "Universities NZ 2025 Target"

list(
  icon = "bullseye",
  color = "warning",
  value = print("70%")
)
```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: "Total OA rate over time"

total |> 
  filter(country == "nz") |> 
  ggplot(aes(x = year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  geom_hline(yintercept = 70, color = "#f39c12", linewidth = 1) +
  annotate("text", x = prev_year - 1, y = 73, label = "UNZ Target", colour = "#f39c12") +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Access type over time

pub_data |> 
  filter(country == "nz") |> 
  group_by(year, key) |> 
  summarise(
    pc = mean(pc),
    count = sum(count)
  ) |> 
  ggplot(aes(x = year, y = pc, fill = key)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

### Column {width=50%}

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pub_data |> 
  filter(country == "nz", year == max(year), key == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc/100,
  ) |> 
  arrange(desc(percent)) |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

# Publications (AU)

#### Row {height=20%}

```{r}
#| content: valuebox
#| title: "Current OA rate"

total_au <- 
  total |> 
  filter(country == "au", year == prev_year) |> 
  pull(pc)

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_au), "%")
)
```

```{r}
#| content: valuebox
#| title: "OA target"

list(
  icon = "bullseye",
  color = "warning",
  value = print("—")
)
```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)
```

#### Row {height=80%}

##### Column {width=50%}

```{r}
#| title: Total OA rate over time

total |> 
  filter(country == "au") |> 
  ggplot(aes(x = year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Access type over time

pub_data |> 
  filter(country == "au") |> 
  group_by(year, key) |> 
  summarise(
    pc = mean(pc),
  ) |> 
  ggplot(aes(x = year, y = pc, fill = key)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours,
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

##### Column {width=50%}

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pub_data |> 
  filter(country == "au", year == max(year), key == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc/100,
  ) |> 
  arrange(desc(percent)) |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

# Comparison

## Row {height=20%}

```{r}
#| content: valuebox
#| title: "Current OA rate (NZ)"

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_nz), "%")
)
```

```{r}
#| content: valuebox
#| title: "Current OA rate (AU)"

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_au), "%")
)
```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=40%}

```{r}
#| title: NZ vs AU OA rate over time

ggplot(total, aes(x = year, y = pc)) +
  geom_line(data = total |> filter(country == "nz"), size = 1, aes(colour = "NZ")) +
  geom_line(data = total |> filter(country == "au"), size = 1, aes(colour = "AU")) +
  theme_classic() +
  scale_color_manual(name = "Country", values = c("NZ" = "#f3f2f2", "AU" = "gold"),
                     guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Institutional trend over time (NZ)

inst_lines_nz <- c("University of Auckland" = "#00467f",
                   "Auckland University of Technology" = "#000000",
                   "University of Waikato" = "#ffa400",
                   "Massey University" = "#8998b9",
                   "Victoria University of Wellington" = "#004b34",
                   "University of Canterbury" = "#ce0818",
                   "Lincoln University" = "#2d4ce5",
                   "University of Otago" = "#f2b043")

inst_lines_au <- c("The University of Melbourne" = "#000c38",
                   "Australian National University" = "#bf872b",
                   "The University of Sydney" = "#141414",
                   "The University of Queensland" = "#411d62",
                   "The University of Western Australia" = "#e2b600",
                   "The University of Adelaide" = "#005a9c",
                   "Monash University" = "#004b34",
                   "UNSW Sydney" = "#da4726")

plot_trend <- 
  function(x) {
    ggplot(x, aes(x = year, y = percent, colour = institution)) +
      geom_line(size = 1) +
      xlab("") +
      ylab("Proportion Open Access (%)") +
      ylim(20, 80) +
      scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
      plot_theme() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
    }
```

# Institutions

```{r}
plot_inst <- function(x) {
  plot_title <- (x$institution) 
  
  plot <- ggplot(x, aes(x = year, y = pc, fill = key)) +
    geom_area() +
    scale_fill_manual(name = "Access Type",
                      values = plot_colours) +
    xlab("") +
    ylab("Proportion (%)") +
    scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
    plot_theme()
  
  plot_final <- plot + ggtitle(plot_title)
  
  return(plot_final)
}

plot_nz_list <- pub_data |>
  filter(country == "nz") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x))

plot_au_list <- pub_data |>
  filter(country == "au") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x))
```

## Column {width=50%}

```{r}
#| title: Institutional Trend (NZ)

pub_data |>
  filter(country == "nz", key == "closed") |>
  summarise(
    institution = institution,
    year = year,
    percent = 100 - pc,
  ) |>
  plot_trend() + 
  scale_colour_manual(values = inst_lines_nz)
```

### Row {.tabset}

```{r}
cat("title=", plot_nz_list[[1]][["data"]][["abbrev"]][1])
plot_nz_list[[1]]
```

```{r}
cat("title=", plot_nz_list[[2]][["data"]][["abbrev"]][1])
plot_nz_list[[2]]
```

```{r}
cat("title=", plot_nz_list[[3]][["data"]][["abbrev"]][1])
plot_nz_list[[3]]
```

```{r}
cat("title=", plot_nz_list[[4]][["data"]][["abbrev"]][1])
plot_nz_list[[4]]
```

```{r}
cat("title=", plot_nz_list[[5]][["data"]][["abbrev"]][1])
plot_nz_list[[5]]
```

```{r}
cat("title=", plot_nz_list[[6]][["data"]][["abbrev"]][1])
plot_nz_list[[6]]
```

```{r}
cat("title=", plot_nz_list[[7]][["data"]][["abbrev"]][1])
plot_nz_list[[7]]
```

```{r}
cat("title=", plot_nz_list[[8]][["data"]][["abbrev"]][1])
plot_nz_list[[8]]
```

## Column {width=50%}

```{r}
#| title: Institutional Trend (AU)

pub_data |>
  filter(country == "au", key == "closed") |>
  summarise(
    institution = institution,
    year = year,
    percent = 100 - pc,
  ) |>
  plot_trend() +
  scale_colour_manual(values = inst_lines_au)
```

### Row {.tabset}

```{r}
cat("title=", plot_au_list[[1]][["data"]][["abbrev"]][1])
plot_au_list[[1]]
```

```{r}
cat("title=", plot_au_list[[2]][["data"]][["abbrev"]][1])
plot_au_list[[2]]
```

```{r}
cat("title=", plot_au_list[[3]][["data"]][["abbrev"]][1])
plot_au_list[[3]]
```

```{r}
cat("title=", plot_au_list[[4]][["data"]][["abbrev"]][1])
plot_au_list[[4]]
```

```{r}
cat("title=", plot_au_list[[5]][["data"]][["abbrev"]][1])
plot_au_list[[5]]
```

```{r}
cat("title=", plot_au_list[[6]][["data"]][["abbrev"]][1])
plot_au_list[[6]]
```

```{r}
cat("title=", plot_au_list[[7]][["data"]][["abbrev"]][1])
plot_au_list[[7]]
```

```{r}
cat("title=", plot_au_list[[8]][["data"]][["abbrev"]][1])
plot_au_list[[8]]
```

# Field (NZ)

```{r}
#| title: "OA Rate By Field (NZ)"

cit_nz |>
  drop_na(field) |> 
  group_by(field, is_oa) |> 
  tally() |> 
  mutate(
    pc = n / sum(n)
    ) |>
  filter(is_oa == TRUE) |>
  ggplot(aes(x = reorder(field, pc), y = pc)) +
  geom_bar(position = "dodge", stat = "identity", fill = "#4DAF4A") +
  geom_text(aes(label = n, y = pc - 0.02),
            vjust = 0.5,
            colour = "white",
            size = 3) +
  coord_flip() +
  plot_theme() +
  ylab("") +
  xlab("") +
  scale_y_continuous(labels = label_percent(),
                     limits = c(0, 0.8)) +
  theme(text = element_text(size = 10))

```

# Field (AU)

```{r}
#| title: "OA Rate By Field (AU)"

cit_au |>
  drop_na(field) |> 
  group_by(field, is_oa) |> 
  tally() |> 
  mutate(
    pc = n / sum(n)
    ) |>
  filter(is_oa == TRUE) |>
  ggplot(aes(x = reorder(field, pc), y = pc)) +
  geom_bar(position = "dodge", stat = "identity", fill = "#4DAF4A") +
  geom_text(aes(label = n, y = pc - 0.02),
            vjust = 0.5,
            colour = "white",
            size = 3) +
  coord_flip() +
  plot_theme() +
  ylab("") +
  xlab("") +
  scale_y_continuous(labels = label_percent(),
                     limits = c(0, 0.8)) +
  theme(text = element_text(size = 10))

```

# Citations (NZ)

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Publications (2020-2024)"

list(
  icon = "file-text",
  color = "info",
  value = format(nrow(cit_nz), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Citations (2020-2024)"

list(
  icon = "blockquote-left",
  color = "info",
  value = format(sum(cit_nz$cited_by_count), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant <-
cit_nz |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

list(
  icon = "bar-chart",
  color = "success",
  value = paste0("+", scales::percent(1-cit_advant$difference))
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/citations-nz.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: "Open vs Closed (2020-2024)"

cit_nz |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme() +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Median Citations")
```


```{r}
#| title: "Open Access Status (2020-2024)"

cit_nz |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme() +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Median Citations")
```

### Column {width=50%}

```{r}
#| title: "Citation Summary (2020-2024)"

cit_nz |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread (2020-2024)"

cit_nz |>
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks = c(0,1,2,5,10,100,1000), trans = "log1p", expand = c(0,0)) +
  plot_theme() +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```

# Citations (AU)

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Publications (2020-2024)"

list(
  icon = "file-text",
  color = "info",
  value = format(nrow(cit_au), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Citations (2020-2024)"

list(
  icon = "blockquote-left",
  color = "info",
  value = format(sum(cit_au$cited_by_count), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant_au <-
cit_au |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

list(
  icon = "bar-chart",
  color = "success",
  value = paste0("+", scales::percent(1-cit_advant_au$difference))
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/citations-au.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: "Open vs Closed (2020-2024)"

cit_au |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme() +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Median Citations")
```


```{r}
#| title: "Open Access Status (2020-2024)"

cit_au |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme() +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Median Citations")
```

### Column {width=50%}

```{r}
#| title: "Citation Summary (2020-2024)"

cit_au |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread (2020-2024)"

cit_au |>
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks = c(0,1,2,5,10,100,1000), trans = "log1p", expand = c(0,0)) +
  plot_theme() +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```


