---
title: "OA Dashboard"
format: 
  dashboard:
    nav-buttons: [github]
    github: https://github.com/tesaunders/oa-nz
    output-file: index.html
logo: images/oa.png
theme:
  - darkly
  - custom.scss
fig-width: 12
---

```{r}
#| label: load-packages
#| message: false

library(dplyr)
library(purrr)
library(ggplot2)
library(ggthemes)
library(gt)
library(tidyr)
library(scales)
library(forcats)
library(stringr)

source("r/functions.r")
```

```{r}
#| label: define-values-themes

prev_year <- as.numeric(format(Sys.Date(), "%Y"))-1

plot_colours <- c("bronze" = "#A65628",
                  "hybrid" = "#377EB8",
                  "gold" = "#FFFF33",
                 "diamond" = "lightgrey",
                 "green" = "#4DAF4A",
                 "closed" = "#E41A1C")

oa_levels <- c("closed", "hybrid", "gold", "green", "diamond", "bronze")

publisher_colours_nz <- c("Elsevier" = "#e43d04",
                          "Wiley" = "black",
                          "Taylor & Francis" = "#181b82",
                          "Springer" = "#102b53",
                          "MDPI" = "#606b84",
                          "Other" = "gold")

publisher_colours_au <- c("Elsevier" = "#e43d04",
                          "Wiley" = "black",
                          "Taylor & Francis" = "#181b82",
                          "Springer" = "#102b53",
                          "Oxford University Press" = "#606b84",
                          "Other" = "gold")

inst_lines_nz <- c("University of Auckland" = "#00467f",
                   "Auckland University of Technology" = "#000000",
                   "University of Waikato" = "#ffa400",
                   "Massey University" = "#8998b9",
                   "Victoria University of Wellington" = "#004b34",
                   "University of Canterbury" = "#ce0818",
                   "Lincoln University" = "#2d4ce5",
                   "University of Otago" = "#f2b043")

inst_lines_au <- c("The University of Melbourne" = "#000c38",
                   "Australian National University" = "#bf872b",
                   "The University of Sydney" = "#141414",
                   "The University of Queensland" = "#411d62",
                   "The University of Western Australia" = "#e2b600",
                   "The University of Adelaide" = "#005a9c",
                   "Monash University" = "#004b34",
                   "UNSW Sydney" = "#da4726")
```

```{r}
#| label: load-data

pubs_all <- readRDS("data/pubs-all.rds")
pubs_summary <- read.csv("data/pubs-summary.csv")

pubs_summary$oa_status <- factor(pubs_summary$oa_status, levels = oa_levels)
```

#  {.sidebar}

This dashboard displays data relating to open access publication output from all 8 Aotearoa New Zealand universities and the Australian 'Group of 8' universities. Data is extracted from [OpenAlex](https://openalex.org/about) via the [API](https://docs.openalex.org/) to examine publication and citation trends since `r min(pubs_all$publication_year)` up to `r prev_year`. Data and code are available on [GitHub](https://github.com/tesaunders/oa-nz/).

------------------------------------------------------------------------

Created and maintained by [Tom Saunders](https://tomsaunders.me).

------------------------------------------------------------------------

::: {.callout-note collapse="true"}
## Notes on data

Data includes all journal articles, books, and book chapters with an author affiliated with at least one of the institutions named above. Works about the publication venue ("paratext") have been excluded (e.g. covers, table of contents, editorial board listing, issue information), as have retracted works.

Works are counted more than once if their authors are affiliated with more than one of the universities examined here. 'Current OA Rates' for NZ and AU universities are calculated by first finding the OA rate for each university for the last full year, and taking the mean of these for each respective region. 

:::

```{r}
#| label: get-open-rates

open_nz <- get_open("NZ")
open_au <- get_open("AU")
```

# Publications

## Row {.tabset}

### NZ

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Current OA rate"

make_valuebox("unlock", "success", percent(mean(open_nz$pc)))
```

```{r}
#| content: valuebox
#| title: "Universities NZ 2025 Target"

make_valuebox("bullseye", "warning", "70%")
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", 
              format(file.info("data/pubs-summary.csv")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: "Total OA rate over time"

annual_open("NZ") |> 
  ggplot(aes(x = publication_year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  geom_hline(yintercept = 0.7, color = "#f39c12", linewidth = 1) +
  annotate("text", x = prev_year - 1, y = 0.73, label = "UNZ Target", colour = "#f39c12") +
  xlab("") +
  ylab("Open Access (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_summary$publication_year)) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark")
```

```{r}
#| title: Access type over time

pubs_summary |> 
  filter(country_code == "NZ") |> 
  group_by(publication_year, oa_status) |> 
  summarise(
    pc = mean(pc),
    count = sum(n)
  ) |> 
  ggplot(aes(x = publication_year, y = pc, fill = oa_status)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_summary$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  plot_theme("dark")
```

:::

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pubs_summary |> 
  filter(country_code == "NZ", 
         publication_year == max(publication_year), 
         oa_status == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc,
  ) |> 
  arrange(desc(percent)) |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::

::::::

### AU

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Current OA rate"

make_valuebox("unlock", "success", percent(mean(open_au$pc)))
```

```{r}
#| content: valuebox
#| title: "OA target"

make_valuebox("bullseye", "warning", print("—"))
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", 
              format(file.info("data/pubs-summary.csv")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: Total OA rate over time

annual_open("AU") |> 
  ggplot(aes(x = publication_year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  xlab("") +
  ylab("Open Access (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_summary$publication_year)) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark")
```

```{r}
#| title: Access type over time

pubs_summary |> 
  filter(country_code == "AU") |> 
  group_by(publication_year, oa_status) |> 
  summarise(
    pc = mean(pc),
    count = sum(n)
  ) |> 
  ggplot(aes(x = publication_year, y = pc, fill = oa_status)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_summary$publication_year)) +
  plot_theme("dark")
```

:::

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pubs_summary |> 
  filter(country_code == "AU", 
         publication_year == max(publication_year), 
         oa_status == "closed") |>
  summarise(
    institution = institution,
    percent = 1- pc,
  ) |> 
  arrange(desc(percent)) |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::

::::::

### Comparison

:::::: {.rows}

```{r}
#| title: NZ vs AU OA rate over time

annual_compare <- full_join(annual_open("NZ"), annual_open("AU")) |> 
  mutate(country_code = case_when(row_number() %in% 1:10 ~ "NZ", .default = "AU"))

ggplot(annual_compare, aes(x = publication_year, y = pc)) +
  geom_line(data = annual_compare |> filter(country_code == "NZ"), 
            size = 1, aes(colour = "NZ")) +
  geom_line(data = annual_compare |> filter(country_code == "AU"), 
            size = 1, aes(colour = "AU")) +
  theme_classic() +
  scale_color_manual(name = "Country", values = c("NZ" = "#f3f2f2", "AU" = "gold"),
                     guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Open Access (%)") +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  scale_x_continuous(n.breaks = prev_year-min(annual_compare$publication_year)) +
  plot_theme("dark")
```

::::::

# Institutions

## Column {width=50%}

```{r}
#| title: Institutional Trend (NZ)

pubs_summary |>
  filter(country_code == "NZ", oa_status == "closed") |>
  reframe(
    institution = institution,
    publication_year = publication_year,
    percent = 1 - pc,
  ) |>
  plot_trend() +
  scale_colour_manual(values = inst_lines_nz) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Row {.tabset}

```{r}
plot_nz_list <- 
  pubs_summary |>
  filter(country_code == "NZ") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x, "dark"))
```

```{r}
cat("title=", plot_nz_list[[1]][["data"]][["abbrev"]][1])
plot_nz_list[[1]]
```

```{r}
cat("title=", plot_nz_list[[2]][["data"]][["abbrev"]][1])
plot_nz_list[[2]]
```

```{r}
cat("title=", plot_nz_list[[3]][["data"]][["abbrev"]][1])
plot_nz_list[[3]]
```

```{r}
cat("title=", plot_nz_list[[4]][["data"]][["abbrev"]][1])
plot_nz_list[[4]]
```

```{r}
cat("title=", plot_nz_list[[5]][["data"]][["abbrev"]][1])
plot_nz_list[[5]]
```

```{r}
cat("title=", plot_nz_list[[6]][["data"]][["abbrev"]][1])
plot_nz_list[[6]]
```

```{r}
cat("title=", plot_nz_list[[7]][["data"]][["abbrev"]][1])
plot_nz_list[[7]]
```

```{r}
cat("title=", plot_nz_list[[8]][["data"]][["abbrev"]][1])
plot_nz_list[[8]]
```

## Column {width=50%}

```{r}
plot_au_list <- 
  pubs_summary |>
  filter(country_code == "AU") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x, "dark"))
```

```{r}
#| title: Institutional Trend (AU)

pubs_summary |>
  filter(country_code == "AU", oa_status == "closed") |>
  reframe(
    institution = institution,
    publication_year = publication_year,
    percent = 1 - pc,
  ) |>
  plot_trend() + 
  scale_colour_manual(values = inst_lines_au) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Row {.tabset}

```{r}
cat("title=", plot_au_list[[1]][["data"]][["abbrev"]][1])
plot_au_list[[1]]
```

```{r}
cat("title=", plot_au_list[[2]][["data"]][["abbrev"]][1])
plot_au_list[[2]]
```

```{r}
cat("title=", plot_au_list[[3]][["data"]][["abbrev"]][1])
plot_au_list[[3]]
```

```{r}
cat("title=", plot_au_list[[4]][["data"]][["abbrev"]][1])
plot_au_list[[4]]
```

```{r}
cat("title=", plot_au_list[[5]][["data"]][["abbrev"]][1])
plot_au_list[[5]]
```

```{r}
cat("title=", plot_au_list[[6]][["data"]][["abbrev"]][1])
plot_au_list[[6]]
```

```{r}
cat("title=", plot_au_list[[7]][["data"]][["abbrev"]][1])
plot_au_list[[7]]
```

```{r}
cat("title=", plot_au_list[[8]][["data"]][["abbrev"]][1])
plot_au_list[[8]]
```

# Field

## Row {.tabset}

### NZ

```{r}
#| title: "OA Rate By Field (NZ)"

pubs_all |>
  filter(country_code == "NZ") |> 
  drop_na(field) |> 
  group_by(field, is_oa) |> 
  tally() |> 
  mutate(
    pc = n / sum(n)
    ) |>
  filter(is_oa == TRUE) |>
  ggplot(aes(x = reorder(field, pc), y = pc)) +
  geom_bar(position = "dodge", stat = "identity", fill = "#4DAF4A") +
  geom_text(aes(label = n, y = pc - 0.02),
            vjust = 0.5,
            colour = "white",
            size = 3) +
  coord_flip() +
  plot_theme("dark") +
  ylab("") +
  xlab("") +
  labs(caption = "Numbers on bars are counts of open works.") +
  scale_y_continuous(labels = label_percent(),
                     limits = c(0, 0.8)) +
  theme(text = element_text(size = 10))
```

### AU

```{r}
#| title: "OA Rate By Field (AU)"

pubs_all |>
  filter(country_code == "AU") |> 
  drop_na(field) |> 
  group_by(field, is_oa) |> 
  tally() |> 
  mutate(
    pc = n / sum(n)
    ) |>
  filter(is_oa == TRUE) |>
  ggplot(aes(x = reorder(field, pc), y = pc)) +
  geom_bar(position = "dodge", stat = "identity", fill = "#4DAF4A") +
  geom_text(aes(label = n, y = pc - 0.02),
            vjust = 0.5,
            colour = "white",
            size = 3) +
  coord_flip() +
  plot_theme("dark") +
  ylab("") +
  xlab("") +
  labs(caption = "Numbers on bars are counts of open works.") +
  scale_y_continuous(labels = label_percent(),
                     limits = c(0, 0.9)) +
  theme(text = element_text(size = 10))
```

# Citations

## Row {.tabset}

### NZ

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Publications"

make_valuebox("file-text", "info", 
              format(nrow(pubs_all |> filter(country_code == "NZ")), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Citations"

make_valuebox("blockquote-left", "info", 
              format(
                pubs_all |> 
                  filter(country_code == "NZ") |> 
                  summarise(citations = sum(cited_by_count)) |> 
                  pull(citations), 
                big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant_nz <-
pubs_all |> 
  filter(country_code == "NZ") |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

make_valuebox("bar-chart", "success", scales::percent(1 - cit_advant_nz$difference))
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", format(file.info("data/pubs-all.csv")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: "Open vs Closed"

pubs_all |>
  filter(country_code == "NZ") |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

```{r}
#| title: "Open Access Status"

pubs_all |>
  filter(country_code == "NZ") |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

:::

::: {.rows}

```{r}
#| title: "Citation Summary"

pubs_all |> 
  filter(country_code == "NZ") |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread"

pubs_all |>
  filter(country_code == "NZ") |> 
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks = c(0,1,2,5,10,100,1000), trans = "log1p", expand = c(0,0)) +
  plot_theme("dark") +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```

:::

::::

::::::

### AU

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Publications"

make_valuebox("file-text", "info", 
              format(nrow(pubs_all |> filter(country_code == "AU")), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Citations"

make_valuebox("blockquote-left", "info", 
              format(
                pubs_all |> 
                  filter(country_code == "AU") |> 
                  summarise(citations = sum(cited_by_count)) |> 
                  pull(citations), 
                big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant_nz <-
pubs_all |> 
  filter(country_code == "AU") |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

make_valuebox("bar-chart", "success", scales::percent(1 - cit_advant_nz$difference))
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", format(file.info("data/pubs-all.csv")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: "Open vs Closed"

pubs_all |>
  filter(country_code == "AU") |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

```{r}
#| title: "Open Access Status"

pubs_all |>
  filter(country_code == "AU") |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

:::

::: {.rows}

```{r}
#| title: "Citation Summary"

pubs_all |> 
  filter(country_code == "AU") |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread"

pubs_all |>
  filter(country_code == "AU") |> 
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks = c(0,1,2,5,10,100,1000), trans = "log1p", expand = c(0,0)) +
  plot_theme("dark") +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```

:::

::::

::::::

# Publishers

## Row {.tabset}

### NZ

:::: {.columns}

::: {.rows}

```{r}
#| label: clean-publishers

pubs_all <-
  pubs_all |> 
  mutate(
    publisher_clean = case_when(
      str_detect(publisher, "Elsevier|RELX|Cell Press") ~ "Elsevier",
      str_detect(publisher, "Wiley") ~ "Wiley",
      str_detect(publisher, "Multidisciplinary Digital Publishing Institute") ~ "MDPI",
      str_detect(publisher, "Taylor & Francis|Routledge|Informa") ~ "Taylor & Francis",
      str_detect(publisher, "Springer|Nature Portfolio") ~ "Springer",
      str_detect(publisher, "Frontiers") ~ "Frontiers Media",
      str_detect(publisher, "IEEE") ~ "IEEE",
      str_detect(publisher, "Institute of Electrical and Electronics Engineers") ~ "IEEE",
      str_detect(publisher, "De Gruyter|Brill") ~ "De Gruyter Brill",
      str_detect(publisher, "Q15088586") ~ "Magnolia Press",
      str_detect(publisher, "Oxford") ~ "Oxford University Press",
      str_detect(publisher, "Wolters|Lippincott") ~ "Wolters Kluwer",
      str_detect(publisher, "Massachusetts Institute of Technology|MIT Press") ~ "The MIT Press",
      .default = publisher)
    )
```

```{r}
#| title: Top 5 Share Over Time

top5 <- 
  pubs_all |> 
  filter(country_code == "NZ", type == "article") |> 
  drop_na(publisher_clean) |> 
  group_by(publisher_clean) |> 
  tally() |> 
  arrange(desc(n)) |> 
  slice_head(n = 5) |> 
  pull(publisher_clean)

pubs_all |> 
  filter(country_code == "NZ", type == "article") |> 
  mutate(
    publisher_clean2 = case_when(
      publisher_clean %in% top5 ~ publisher_clean,
      .default = "Other"
    )
  ) |> 
  group_by(publication_year, publisher_clean2) |> 
  tally() |> 
  mutate(
    pc = n / sum(n),
  ) |> 
  select(-n) |>
  ggplot(aes(x = publication_year, y = pc, fill = publisher_clean2)) +
  geom_area() +
  xlab("") +
  ylab("Article Share (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_nz) +
  plot_theme("dark")
```

```{r}
#| title: Top 5 By Field

pubs_all |>
  filter(country_code == "NZ", type == "article") |>
  drop_na(field) |>
  filter(publisher_clean %in% top5) |>
  group_by(publisher_clean, field) |>
  summarise(
    n = n()
  ) |>
  mutate(
    pc = n/ sum(n)
  ) |>
  ggplot(aes(x = field, y = pc, fill = publisher_clean)) +
  geom_bar(position = "fill", stat = "identity") +
  coord_flip() +
  xlab("") +
  ylab("") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_nz[1:5]) +
  plot_theme("dark")
```

:::

```{r}
#| title: Top 20 by Article Share

publisher_rates <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "NZ", type == "article") |> 
  group_by(publisher_clean, is_oa) |> 
  tally() |> 
  mutate(
    percent_open = n / sum(n)
  ) |> 
  filter(is_oa == TRUE) |> 
  arrange(desc(n))

publisher_citations <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "NZ", type == "article") |> 
  group_by(publisher_clean) |> 
  summarise(
    citations = sum(cited_by_count),
    articles = n(),
    median_citations = median(cited_by_count)
  ) |> 
  filter(articles > 20) |> 
  arrange(desc(median_citations))

nz_articles <- nrow(filter(pubs_all, country_code == "NZ", type == "article"))

left_join(publisher_rates, publisher_citations) |> 
  rename(Publisher = publisher_clean,
         Articles = articles,
         "Percent OA" = percent_open,
         "Median Citations" = median_citations) |> 
  select(-is_oa, -n, -citations) |> 
  relocate(c(Articles, "Percent OA", "Median Citations"), .after = Publisher) |>
  arrange(desc(Articles)) |>
  mutate(
    "Article Share" = Articles / nz_articles
  ) |> 
  ungroup() |> 
  slice_head(n = 20) |> 
  mutate(
    Rank = row_number()
  ) |>
  relocate(Rank) |> 
  relocate("Article Share", .after = Articles) |> 
  gt() |> 
  fmt_number(columns = c(Articles, "Median Citations"), 
             sep_mark = ",", decimals = 0) |> 
  fmt_percent(columns = c("Percent OA", "Article Share"), decimals = 0) |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::

### AU

:::: {.columns}

::: {.rows}

```{r}
#| title: Top 5 Share Over Time

top5 <- 
  pubs_all |> 
  filter(country_code == "AU", type == "article") |> 
  drop_na(publisher_clean) |> 
  group_by(publisher_clean) |> 
  tally() |> 
  arrange(desc(n)) |> 
  slice_head(n = 5) |> 
  pull(publisher_clean)

pubs_all |> 
  filter(country_code == "AU", type == "article") |> 
  mutate(
    publisher_clean2 = case_when(
      publisher_clean %in% top5 ~ publisher_clean,
      .default = "Other"
    )
  ) |> 
  group_by(publication_year, publisher_clean2) |> 
  tally() |> 
  mutate(
    pc = n / sum(n),
  ) |> 
  select(-n) |>
  ggplot(aes(x = publication_year, y = pc, fill = publisher_clean2)) +
  geom_area() +
  xlab("") +
  ylab("Article Share (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_au) +
  plot_theme("dark")
```

```{r}
#| title: Top 5 By Field

pubs_all |>
  filter(country_code == "AU", type == "article") |>
  drop_na(field) |>
  filter(publisher_clean %in% top5) |>
  group_by(publisher_clean, field) |>
  summarise(
    n = n()
  ) |>
  mutate(
    pc = n/ sum(n)
  ) |>
  ggplot(aes(x = field, y = pc, fill = publisher_clean)) +
  geom_bar(position = "fill", stat = "identity") +
  coord_flip() +
  xlab("") +
  ylab("") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_au[1:5]) +
  plot_theme("dark")
```

:::

```{r}
#| title: Top 20 by Article Share

publisher_rates <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "AU", type == "article") |> 
  group_by(publisher_clean, is_oa) |> 
  tally() |> 
  mutate(
    percent_open = n / sum(n)
  ) |> 
  filter(is_oa == TRUE) |> 
  arrange(desc(n))

publisher_citations <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "AU", type == "article") |> 
  group_by(publisher_clean) |> 
  summarise(
    citations = sum(cited_by_count),
    articles = n(),
    median_citations = median(cited_by_count)
  ) |> 
  filter(articles > 20) |> 
  arrange(desc(median_citations))

nz_articles <- nrow(filter(pubs_all, country_code == "AU", type == "article"))

left_join(publisher_rates, publisher_citations) |> 
  rename(Publisher = publisher_clean,
         Articles = articles,
         "Percent OA" = percent_open,
         "Median Citations" = median_citations) |> 
  select(-is_oa, -n, -citations) |> 
  relocate(c(Articles, "Percent OA", "Median Citations"), .after = Publisher) |>
  arrange(desc(Articles)) |>
  mutate(
    "Article Share" = Articles / nz_articles
  ) |> 
  ungroup() |> 
  slice_head(n = 20) |> 
  mutate(
    Rank = row_number()
  ) |>
  relocate(Rank) |> 
  relocate("Article Share", .after = Articles) |> 
  gt() |> 
  fmt_number(columns = c(Articles, "Median Citations"), 
             sep_mark = ",", decimals = 0) |> 
  fmt_percent(columns = c("Percent OA", "Article Share"), decimals = 0) |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::
