---
title: "OA Dashboard"
format: 
  dashboard:
    nav-buttons: [github]
    github: https://github.com/tesaunders/oa-nz
    output-file: index.html
logo: images/oa.png
theme:
  - darkly
  - custom.scss
fig-width: 12
---

```{r}
#| label: load-packages
#| message: false

library(dplyr)
library(purrr)
library(ggplot2)
library(ggthemes)
library(DT)
library(tidyr)
library(scales)

theme_set(theme_minimal(base_size = 24, base_family = "Atkinson Hyperlegible"))
```

```{r}
#| label: load-data
#| message: false

pub_data <- read.csv("data/publications.csv")
cit_data <- read.csv("data/citations.csv")
```

```{r}
#| label: define-values
#| results: hide

prev_year <- as.numeric(format(Sys.Date(), "%Y"))-1
```

```{r}
#| label: define-plot-theme
#| results: hide

plot_theme <- function(base_size=20, base_family="sans") {
   library(ggthemes)
   (theme_foundation(base_size=base_size, base_family=base_family)
      + theme(plot.title = element_text(face = "bold", colour = '#ffffb3',
                                        size = rel(1.2), hjust = 0.5, margin = margin(0,0,20,0)),
              text = element_text(),
              panel.background = element_rect(colour = NA, fill = 'grey20'),
              plot.background = element_rect(colour = NA, fill = '#2d2d2d'),
              panel.border = element_rect(colour = NA),
              axis.title = element_text(face = "bold",size = rel(1), colour = 'white'),
              axis.title.y = element_text(angle=90,vjust =2),
              axis.title.x = element_text(vjust = -0.2),
              axis.text = element_text(colour = 'grey70'), 
              axis.line.x = element_line(colour="grey70"),
              axis.line.y = element_line(colour="grey70"),
              axis.ticks = element_line(colour="grey70"),
              panel.grid.major = element_line(colour="#262626"),
              panel.grid.minor = element_blank(),
              legend.background = element_rect(fill ='#262626'),
              legend.text = element_text(color = 'white'),
              legend.key = element_rect(colour = NA, fill = '#262626'),
              legend.position = "right",
              legend.direction = "vertical",
              legend.box = "vertical",
              legend.key.size= unit(0.5, "cm"),
              #legend.margin = unit(0, "cm"),
              legend.title = element_blank(),
              plot.margin=unit(c(10,5,5,5),"mm"),
              strip.background=element_rect(colour="#2D3A4C",fill="#2D3A4C"),
              strip.text = element_text(face="bold", colour = 'white')
      ))
}

scale_oa <- scale_colour_manual(name = "Access Type",
                      values = c("bronze" = "#A65628",
                                 "hybrid" = "#377EB8",
                                 "gold" = "#FFFF33",
                                 "diamond" = "lightgrey",
                                 "green" = "#4DAF4A",
                                 "closed" = "#E41A1C"))

```

#  {.sidebar}

This dashboard displays data relating to open access publication output from all 8 Aotearoa New Zealand universities and the Australian 'Group of 8' universities.

------------------------------------------------------------------------

Data is extracted from [OpenAlex](https://openalex.org/about) via the [API](https://docs.openalex.org/) to examine publication trends since 2010. Currently, data is only included up to the most recent full year (`r prev_year`). Data and code are available on [GitHub](https://github.com/tesaunders/oa-nz/).

------------------------------------------------------------------------

Created and maintained by [Tom Saunders](https://tomsaunders.me).

------------------------------------------------------------------------

::: {.callout-note collapse="true"}
## Notes on data

Data includes all journal articles, books, and book chapters with an author affiliated with at least one of the institutions named above. Works about the publication venue ("paratext") have been excluded (e.g. covers, table of contents, editorial board listing, issue information), as have retracted works.

:::

```{r}
#| label: summarise-data
#| message: false

total <- pub_data |> 
  mutate(
    open = case_when(key != "closed" ~ "open", .default = key),
  ) |> 
  filter(open == "open") |> 
  group_by(year, country, institution) |> 
  summarise(
    pc = sum(pc),
  ) |> 
  group_by(year, country) |> 
  summarise(
    pc = mean(pc),
  )

```

# New Zealand

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Current OA rate"
#| message: false

total_nz <- total |> 
  filter(country == "nz", year == prev_year) |> 
  pull(pc)

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_nz), "%")
)

```

```{r}
#| content: valuebox
#| title: "Universities NZ 2025 Target"
#| message: false

list(
  icon = "bullseye",
  color = "warning",
  value = print("70%")
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"
#| message: false

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)

```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: Total OA rate over time

trend_nz <- total |> 
  filter(country == "nz")

ggplot(trend_nz, aes(x = year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  geom_hline(yintercept = 70, color = "#f39c12", linewidth = 1) +
  annotate("text", x = prev_year - 1, y = 73, label = "UNZ Target", colour = "#f39c12") +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Access type over time

type_nz <- pub_data |> 
  filter(country == "nz") |> 
  group_by(year, key) |> 
  summarise(
    pc = mean(pc),
  )

ggplot(type_nz, aes(x = year, y = pc, colour = key)) +
  geom_line(size = 1) +
  scale_oa +
  xlab("") +
  ylab("Proportion (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

### Column {width=50%}

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
recent_nz <- pub_data |> 
  filter(country == "nz", year == max(year), key == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc/100,
  ) |> 
  arrange(desc(percent))

recent_nz |> 
  datatable(options = list(dom = 't', ordering = FALSE), rownames = FALSE, colnames=NULL) |> 
  formatPercentage("percent")
```

# Australia G8

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Current OA rate"

total_au <- total |> 
  filter(country == "au", year == prev_year) |> 
  pull(pc)

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_au), "%")
)

```

```{r}
#| content: valuebox
#| title: "OA target"

list(
  icon = "bullseye",
  color = "warning",
  value = print("—")
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)

```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: Total OA rate over time

trend_au <- total |> 
  filter(country == "au")

ggplot(trend_au, aes(x = year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Access type over time

type_au <- pub_data |> 
  filter(country == "au") |> 
  group_by(year, key) |> 
  summarise(
    pc = mean(pc),
  )

ggplot(type_au, aes(x = year, y = pc, colour = key)) +
  geom_line(size = 1) +
  scale_oa +
  xlab("") +
  ylab("Proportion (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

### Column {width=50%}

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
recent_au <- pub_data |> 
  filter(country == "au", year == max(year), key == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc/100,
  ) |> 
  arrange(desc(percent))

recent_au |> 
  datatable(options = list(dom = 't', ordering = FALSE), rownames = FALSE, colnames = NULL) |> 
  formatPercentage("percent")

```

# Comparison

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Current OA rate (NZ)"
#| message: false

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_nz), "%")
)

```

```{r}
#| content: valuebox
#| title: "Current OA rate (AU)"
#| message: false

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_au), "%")
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"
#| message: false

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)

```

## Row {height="40%"}

```{r}
#| title: NZ vs AU OA rate over time

ggplot(total, aes(x = year, y = pc)) +
  geom_line(data = total |> filter(country == "nz"), size = 1, aes(colour = "NZ")) +
  geom_line(data = total |> filter(country == "au"), size = 1, aes(colour = "AU")) +
  theme_classic() +
  scale_color_manual(name = "Country",
                     values = c("NZ" = "#f3f2f2",
                                "AU" = "gold")) +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

## Row {height="40%"}

```{r}
#| title: Institutional trend over time (NZ)

unis_nz <- pub_data |>
  filter(country == "nz", key == "closed") |>
  summarise(
    institution = institution,
    year = year,
    percent = 100 - pc,
  )

ggplot(unis_nz, aes(x = year, y = percent, colour = institution)) +
  geom_line(size = 1) +
  scale_colour_manual(name = "Institution",
                      values = c("University of Auckland" = "#00467f",
                                 "Auckland University of Technology" = "#000000",
                                 "University of Waikato" = "#ffa400",
                                 "Massey University" = "#8998b9",
                                 "Victoria University of Wellington" = "#004b34",
                                 "University of Canterbury" = "#ce0818",
                                 "Lincoln University" = "#2d4ce5",
                                 "University of Otago" = "#f2b043")) +
  theme_classic() +
  xlab("") +
  ylab("Proportion Open Access (%)") +
  ylim(20, 70) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r}
#| title: Institutional trend over time (AU)

unis_au <- pub_data |>
  filter(country == "au", key == "closed") |>
  summarise(
    institution = institution,
    year = year,
    percent = 100 - pc,
  )

ggplot(unis_au, aes(x = year, y = percent, colour = institution)) +
  geom_line(size = 1) +
  scale_colour_manual(name = "Institution",
                      values = c("University of Melbourne" = "#000c38",
                                 "Australian National University" = "#bf872b",
                                 "University of Sydney" = "#141414",
                                 "University of Queensland" = "#411d62",
                                 "University of Western Australia" = "#e2b600",
                                 "University of Adelaide" = "#005a9c",
                                 "Monash University" = "#004b34",
                                 "UNSW Sydney" = "#da4726")) +
  theme_classic() +
  xlab("") +
  ylab("Proportion Open Access (%)") +
  ylim(20, 70) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

# Institutions 

```{r}
plot_inst <- function(pub_data) {
  p <- ggplot(pub_data, aes(x = year, y = pc, colour = key)) +
    geom_line(size = 1) +
    scale_oa +
    xlab("") +
    ylab("Proportion (%)") +
    ylim(0, 80) +
    scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
    ggtitle(paste0(unique(pub_data$institution), 
                   ", ", min(pub_data$year), "-", prev_year)) +
    plot_theme()
  return(p)
}

plot_nz_list <- pub_data |>
  filter(country == "nz") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x))

plot_au_list <- pub_data |>
  filter(country == "au") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x))
```

## Row {.tabset}

```{r}
cat("title=", sort(unique(plot_nz_list[[1]][["data"]][["institution"]])))
plot_nz_list[[1]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[2]][["data"]][["institution"]])))
plot_nz_list[[2]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[3]][["data"]][["institution"]])))
plot_nz_list[[3]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[4]][["data"]][["institution"]])))
plot_nz_list[[4]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[5]][["data"]][["institution"]])))
plot_nz_list[[5]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[6]][["data"]][["institution"]])))
plot_nz_list[[6]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[7]][["data"]][["institution"]])))
plot_nz_list[[7]]
```

```{r}
cat("title=", sort(unique(plot_nz_list[[8]][["data"]][["institution"]])))
plot_nz_list[[8]]
```

## Row {.tabset}

```{r}
cat("title=", sort(unique(plot_au_list[[1]][["data"]][["institution"]])))
plot_au_list[[1]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[2]][["data"]][["institution"]])))
plot_au_list[[2]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[3]][["data"]][["institution"]])))
plot_au_list[[3]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[4]][["data"]][["institution"]])))
plot_au_list[[4]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[5]][["data"]][["institution"]])))
plot_au_list[[5]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[6]][["data"]][["institution"]])))
plot_au_list[[6]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[7]][["data"]][["institution"]])))
plot_au_list[[7]]
```

```{r}
cat("title=", sort(unique(plot_au_list[[8]][["data"]][["institution"]])))
plot_au_list[[8]]
```

# Citations (NZ)

## Row {height="20%"}

```{r}
#| content: valuebox
#| message: false
#| title: "Publications (2020-2024)"

# cat("#| title:",'"',"Total Publications", paste0("(",min(cit_data$publication_year),"-",max(cit_data$publication_year),")", '"'))

list(
  icon = "file-text",
  color = "info",
  value = format(nrow(cit_data), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| message: false
#| title: "Citations (2020-2024)"

# cat("#| title:",'"',"Total Citations", paste0("(",min(cit_data$publication_year),"-",max(cit_data$publication_year),")", '"'))

list(
  icon = "blockquote-left",
  color = "info",
  value = format(sum(cit_data$cited_by_count), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"
#| message: false

cit_advant <-
cit_data |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

list(
  icon = "bar-chart",
  color = "success",
  value = paste0("+", scales::percent(1-cit_advant$difference))
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"
#| message: false

list(
  color = "light",
  value = format(file.info("data/citations.csv")$mtime, format = "%d %b %Y")
)

```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: "Total Citations (Works Published 2020-2024)"

cit_data |> 
  group_by(publication_year) |> 
  summarise(
    total = sum(cited_by_count)
  ) |> 
  arrange(desc(publication_year)) |> 
  ggplot(aes(x = publication_year, y = total)) +
  geom_col(fill = "#f3f2f2") +
  plot_theme() +
  xlab("") +
  ylab("") +
  scale_y_continuous(labels = comma)

```

```{r}
#| title: "Citation Spread (Works Published 2020-2024)"

cit_data |>
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y=..count..), color="black", fill="#f3f2f2") +
  scale_x_continuous(breaks=c(0,1,2,3,4,5,10), trans="log1p", expand=c(0,0)) +
  plot_theme() +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)

```

### Column {width=50%}

```{r}
#| title: "Citation Summary (Works Published 2020-2024)"
 
cit_data |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |> 
  rename(Year = publication_year) |> 
  datatable(options = list(dom = 't', ordering = FALSE), rownames = FALSE)

```


