---
title: "OA Dashboard"
format: 
  dashboard:
    nav-buttons: [github]
    github: https://github.com/tesaunders/oa-nz
    output-file: index.html
logo: images/oa.png
theme:
  - darkly
fig-width: 12
---

```{r}
#| label: load-packages
#| message: false

library(dplyr)
library(purrr)
library(ggplot2)
library(ggthemes)
library(gt)
library(tidyr)
library(scales)
library(forcats)
```

```{r}
#| label: define-values-themes

prev_year <- as.numeric(format(Sys.Date(), "%Y"))-1

plot_theme <- function(base_size=20, base_family="sans") {
   library(ggthemes)
   (theme_foundation(base_size=base_size, base_family=base_family)
      + theme(plot.title = element_text(face = "bold", colour = '#ffffb3',
                                        size = rel(1.2), hjust = 0.5, margin = margin(0,0,20,0)),
              text = element_text(),
              panel.background = element_rect(colour = NA, fill = 'grey20'),
              plot.background = element_rect(colour = NA, fill = '#2d2d2d'),
              panel.border = element_rect(colour = NA),
              axis.title = element_text(face = "bold",size = rel(1), colour = 'white'),
              axis.title.y = element_text(angle=90,vjust =2),
              axis.title.x = element_text(vjust = -0.2),
              axis.text = element_text(colour = 'grey70'), 
              axis.line.x = element_line(colour="grey70"),
              axis.line.y = element_line(colour="grey70"),
              axis.ticks = element_line(colour="grey70"),
              panel.grid.major = element_line(colour="#262626"),
              panel.grid.minor = element_blank(),
              legend.background = element_rect(fill ='#262626'),
              legend.text = element_text(color = 'white'),
              legend.key = element_rect(colour = NA, fill = '#262626'),
              legend.position = "right",
              legend.direction = "vertical",
              legend.box = "vertical",
              legend.key.size= unit(0.5, "cm"),
              #legend.margin = unit(0, "cm"),
              legend.title = element_blank(),
              plot.margin=unit(c(10,5,5,5),"mm"),
              strip.background=element_rect(colour="#2D3A4C",fill="#2D3A4C"),
              strip.text = element_text(face="bold", colour = 'white')
      ))
}

plot_colours <- c("bronze" = "#A65628",
                  "hybrid" = "#377EB8",
                  "gold" = "#FFFF33",
                 "diamond" = "lightgrey",
                 "green" = "#4DAF4A",
                 "closed" = "#E41A1C")
```

```{r}
#| label: load-data

pub_data <- read.csv("data/publications.csv")
cit_data <- read.csv("data/citations.csv")

pub_data$key <- factor(pub_data$key, levels = c("closed", "hybrid", "gold", "green", "diamond", "bronze"))
pub_data$institution <- factor(pub_data$institution)

cit_data$oa_status <- factor(cit_data$oa_status, levels = c("closed", "hybrid", "gold", "green", "diamond", "bronze"))
```

#  {.sidebar}

This dashboard displays data relating to open access publication output from all 8 Aotearoa New Zealand universities and the Australian 'Group of 8' universities.

------------------------------------------------------------------------

Data is extracted from [OpenAlex](https://openalex.org/about) via the [API](https://docs.openalex.org/) to examine publication trends since `r min(pub_data$year)`. Currently, data is only included up to the most recent full year (`r prev_year`). Data and code are available on [GitHub](https://github.com/tesaunders/oa-nz/).

------------------------------------------------------------------------

Created and maintained by [Tom Saunders](https://tomsaunders.me).

------------------------------------------------------------------------

::: {.callout-note collapse="true"}
## Notes on data

Data includes all journal articles, books, and book chapters with an author affiliated with at least one of the institutions named above. Works about the publication venue ("paratext") have been excluded (e.g. covers, table of contents, editorial board listing, issue information), as have retracted works.

:::

```{r}
#| label: summarise-data

total <- 
  pub_data |> 
  mutate(
    open = case_when(key != "closed" ~ "open", .default = key),
  ) |> 
  filter(open == "open") |> 
  group_by(year, country, institution) |> 
  summarise(
    pc = sum(pc),
  ) |> 
  group_by(year, country) |> 
  summarise(
    pc = mean(pc),
  )
```

# New Zealand

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Current OA rate"

total_nz <- 
  total |> 
  filter(country == "nz", year == prev_year) |> 
  pull(pc)

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_nz), "%")
)
```

```{r}
#| content: valuebox
#| title: "Universities NZ 2025 Target"

list(
  icon = "bullseye",
  color = "warning",
  value = print("70%")
)
```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: "Total OA rate over time"

total |> 
  filter(country == "nz") |> 
  ggplot(aes(x = year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  geom_hline(yintercept = 70, color = "#f39c12", linewidth = 1) +
  annotate("text", x = prev_year - 1, y = 73, label = "UNZ Target", colour = "#f39c12") +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Access type over time (Combined)

pub_data |> 
  filter(country == "nz") |> 
  group_by(year, key) |> 
  summarise(
    pc = mean(pc),
    count = sum(count)
  ) |> 
  ggplot(aes(x = year, y = pc, fill = key)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

### Column {width=50%}

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pub_data |> 
  filter(country == "nz", year == max(year), key == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc/100,
  ) |> 
  arrange(desc(percent)) |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

# Australia G8

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Current OA rate"

total_au <- 
  total |> 
  filter(country == "au", year == prev_year) |> 
  pull(pc)

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_au), "%")
)
```

```{r}
#| content: valuebox
#| title: "OA target"

list(
  icon = "bullseye",
  color = "warning",
  value = print("—")
)
```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: Total OA rate over time

total |> 
  filter(country == "au") |> 
  ggplot(aes(x = year, y = pc)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Access type over time

pub_data |> 
  filter(country == "au") |> 
  group_by(year, key) |> 
  summarise(
    pc = mean(pc),
  ) |> 
  ggplot(aes(x = year, y = pc, fill = key)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours,
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

### Column {width=50%}

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pub_data |> 
  filter(country == "au", year == max(year), key == "closed") |>
  summarise(
    institution = institution,
    percent = 1 - pc/100,
  ) |> 
  arrange(desc(percent)) |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

# Comparison

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Current OA rate (NZ)"

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_nz), "%")
)
```

```{r}
#| content: valuebox
#| title: "Current OA rate (AU)"

list(
  icon = "unlock",
  color = "success",
  value = paste0(round(total_au), "%")
)
```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/publications.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=40%}

```{r}
#| title: NZ vs AU OA rate over time

ggplot(total, aes(x = year, y = pc)) +
  geom_line(data = total |> filter(country == "nz"), size = 1, aes(colour = "NZ")) +
  geom_line(data = total |> filter(country == "au"), size = 1, aes(colour = "AU")) +
  theme_classic() +
  scale_color_manual(name = "Country", values = c("NZ" = "#f3f2f2", "AU" = "gold"),
                     guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Open Access (%)") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
  plot_theme()
```

```{r}
#| title: Institutional trend over time (NZ)

inst_lines_nz <- c("University of Auckland" = "#00467f",
                   "Auckland University of Technology" = "#000000",
                   "University of Waikato" = "#ffa400",
                   "Massey University" = "#8998b9",
                   "Victoria University of Wellington" = "#004b34",
                   "University of Canterbury" = "#ce0818",
                   "Lincoln University" = "#2d4ce5",
                   "University of Otago" = "#f2b043")

inst_lines_au <- c("University of Melbourne" = "#000c38",
                   "Australian National University" = "#bf872b",
                   "University of Sydney" = "#141414",
                   "University of Queensland" = "#411d62",
                   "University of Western Australia" = "#e2b600",
                   "University of Adelaide" = "#005a9c",
                   "Monash University" = "#004b34",
                   "UNSW Sydney" = "#da4726")

plot_trend <- 
  function(x) {
    ggplot(x, aes(x = year, y = percent, colour = institution)) +
      geom_line(size = 1) +
      xlab("") +
      ylab("Proportion Open Access (%)") +
      ylim(20, 70) +
      scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
      plot_theme() +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
    }
```

## Row {height=40%}

```{r}
#| title: Institutional Trend (NZ)

pub_data |>
  filter(country == "nz", key == "closed") |>
  summarise(
    institution = institution,
    year = year,
    percent = 100 - pc,
  ) |>
  plot_trend() + 
  scale_colour_manual(name = "Institution", values = inst_lines_nz)
```


```{r}
#| title: Institutional Trend (AU)

pub_data |>
  filter(country == "au", key == "closed") |>
  summarise(
    institution = institution,
    year = year,
    percent = 100 - pc,
  ) |>
  plot_trend() +
  scale_colour_manual(name = "Institution", values = inst_lines_au)
```

# Institutions 

```{r}
plot_inst <- function(x) {
    ggplot(x, aes(x = year, y = pc, fill = key)) +
    geom_area() +
    scale_fill_manual(name = "Access Type",
                      values = plot_colours) +
    xlab("") +
    ylab("Proportion (%)") +
    scale_x_continuous(n.breaks = prev_year-min(pub_data$year)) +
    plot_theme()
}

plot_nz_list <- pub_data |>
  filter(country == "nz") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x))

plot_titles_nz <- 
  pub_data |> 
  filter(country == "nz") |> 
  group_by(institution) |> 
  group_keys() |> 
  pull(institution) |> 
  as.character()

plot_au_list <- pub_data |>
  filter(country == "au") |> 
  group_by(institution, .add = TRUE) |> 
  group_split(institution) |> 
  map(~ plot_inst(.x))

plot_titles_au <- 
  pub_data |> 
  filter(country == "au") |> 
  group_by(institution) |> 
  group_keys() |> 
  pull(institution) |> 
  as.character()
```

## Row {.tabset}

```{r}
cat("title=", plot_titles_nz[1])
plot_nz_list[[1]]
```

```{r}
cat("title=", plot_titles_nz[2])
plot_nz_list[[2]]
```

```{r}
cat("title=", plot_titles_nz[3])
plot_nz_list[[3]]
```

```{r}
cat("title=", plot_titles_nz[4])
plot_nz_list[[4]]
```

```{r}
cat("title=", plot_titles_nz[5])
plot_nz_list[[5]]
```

```{r}
cat("title=", plot_titles_nz[6])
plot_nz_list[[6]]
```

```{r}
cat("title=", plot_titles_nz[7])
plot_nz_list[[7]]
```

```{r}
cat("title=", plot_titles_nz[8])
plot_nz_list[[8]]
```

## Row {.tabset}

```{r}
cat("title=", plot_titles_au[1])
plot_au_list[[1]]
```

```{r}
cat("title=", plot_titles_au[2])
plot_au_list[[2]]
```

```{r}
cat("title=", plot_titles_au[3])
plot_au_list[[3]]
```

```{r}
cat("title=", plot_titles_au[4])
plot_au_list[[4]]
```

```{r}
cat("title=", plot_titles_au[5])
plot_au_list[[5]]
```

```{r}
cat("title=", plot_titles_au[6])
plot_au_list[[6]]
```

```{r}
cat("title=", plot_titles_au[7])
plot_au_list[[7]]
```

```{r}
cat("title=", plot_titles_au[8])
plot_au_list[[8]]
```

# Citations (NZ)

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Publications (2020-2024)"

list(
  icon = "file-text",
  color = "info",
  value = format(nrow(cit_data), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Citations (2020-2024)"

list(
  icon = "blockquote-left",
  color = "info",
  value = format(sum(cit_data$cited_by_count), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant <-
cit_data |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

list(
  icon = "bar-chart",
  color = "success",
  value = paste0("+", scales::percent(1-cit_advant$difference))
)

```

```{r}
#| content: valuebox
#| title: "Last Updated"

list(
  color = "light",
  value = format(file.info("data/citations.csv")$mtime, format = "%d %b %Y")
)
```

## Row {height=80%}

### Column {width=50%}

```{r}
#| title: "Median Citations: Open vs Closed (Works Published 2020-2024)"

cit_data |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme() +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Median Citations")
```


```{r}
#| title: "Median Citations: Open Access Status (Works Published 2020-2024)"

cit_data |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme() +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Median Citations")
```

### Column {width=50%}

```{r}
#| title: "Citation Summary (Works Published 2020-2024)"

cit_data |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread (Works Published 2020-2024)"

cit_data |>
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks=c(0,1,2,3,4,5,10), trans = "log1p", expand = c(0,0)) +
  plot_theme() +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```
