---
title: "OA Dashboard"
format: 
  dashboard:
    nav-buttons: [github]
    github: https://github.com/tesaunders/oa-nz
    output-file: index.html
logo: images/oa.png
theme:
  - darkly
  - custom.scss
fig-width: 12
---

```{r setup, include = FALSE}
library(dplyr)
library(purrr)
library(ggplot2)
library(gt)
library(tidyr)
library(scales)
library(forcats)
library(stringr)

source("r/functions.r")
source("r/objects.r")
```

```{r load-data}
pubs_all <- readRDS("data/pubs-all.rds")
pubs_all$oa_status <- factor(pubs_all$oa_status, c("closed", "hybrid", "gold", "green", "diamond", "bronze"))
```

#  {.sidebar}

This dashboard displays data relating to open access publication output from all 8 Aotearoa New Zealand universities and the Australian 'Group of 8' universities. Data is extracted from [OpenAlex](https://openalex.org/about) via the [API](https://docs.openalex.org/) to examine publication and citation trends since `r min(pubs_all$publication_year)` up to `r prev_year`. Data and code are available on [GitHub](https://github.com/tesaunders/oa-nz/).

------------------------------------------------------------------------

Created and maintained by [Tom Saunders](https://tomsaunders.me).

------------------------------------------------------------------------

::: {.callout-note collapse="true"}
## Notes on data

Data includes all journal articles, books, and book chapters with an author affiliated with at least one of the institutions named above. Works about the publication venue ("paratext") have been excluded (e.g. covers, table of contents, editorial board listing, issue information), as have retracted works.

*Current open access rates:

- count each work only once, whereas open access rates by institution will count a work under multiple institutions if those institutions are represented in the author affiliations. 
- are based on works published in the previous full year.

:::

```{r get-open-rates}
open_overall_nz <- get_overall_open("NZ")
open_overall_au <- get_overall_open("AU")
```

# Publications

## Row {.tabset}

### NZ

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Current OA rate"

make_valuebox("unlock", "success", paste0(percent(open_overall_nz), "*"))
```

```{r}
#| content: valuebox
#| title: "Universities NZ 2025 Target"

make_valuebox("bullseye", "warning", "70%")
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", 
              format(file.info("data/pubs-all.rds")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: "Total OA rate over time"

get_annual_open("NZ") |> 
  ggplot(aes(x = publication_year, y = freq)) +
  geom_line(linewidth = 1, colour = "#f3f2f2") +
  geom_hline(yintercept = 0.7, color = "#f39c12", linewidth = 1) +
  annotate("text", x = prev_year - 1, y = 0.73, label = "UNZ Target", colour = "#f39c12") +
  xlab("") +
  ylab("Open Access (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark")
```

```{r}
#| title: Access type over time

pubs_all |>
  filter(country_code == "NZ") |>
  group_by(item_id) |>
  mutate(duplicate = n()) |>
  filter(duplicate == 1) |>
  group_by(oa_status, publication_year) |>
  summarise(n = n()) |>
  group_by(publication_year) |>
  mutate(freq = n / sum(n)) |>
  ggplot(aes(x = publication_year, y = freq, fill = oa_status)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  plot_theme("dark")
```

:::

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pubs_all |> 
  filter(country_code == "NZ", 
         publication_year == max(publication_year)) |>
  group_by(institution, is_oa) |> 
  tally() |> 
  pivot_wider(names_from = is_oa, values_from = n) |> 
  mutate(pc_open = `TRUE` / (`TRUE` + `FALSE`)) |> 
  arrange(desc(pc_open)) |> 
  select(-`FALSE`, -`TRUE`) |>
  ungroup() |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::

::::::

### AU

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Current OA rate"

make_valuebox("unlock", "success", paste0(percent(open_overall_au), "*"))
```

```{r}
#| content: valuebox
#| title: "OA target"

make_valuebox("bullseye", "warning", print("—"))
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", 
              format(file.info("data/pubs-all.rds")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: Total OA rate over time

get_annual_open("AU") |> 
  ggplot(aes(x = publication_year, y = freq)) +
  geom_line(size = 1, colour = "#f3f2f2") +
  xlab("") +
  ylab("Open Access (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark")
```

```{r}
#| title: Access type over time

pubs_all |>
  filter(country_code == "AU") |>
  group_by(item_id) |>
  mutate(duplicate = n()) |>
  filter(duplicate == 1) |>
  group_by(oa_status, publication_year) |>
  summarise(n = n()) |>
  group_by(publication_year) |>
  mutate(freq = n / sum(n)) |>
  ggplot(aes(x = publication_year, y = freq, fill = oa_status)) +
  geom_area() +
  scale_fill_manual(name = "Access Type", values = plot_colours) +
  xlab("") +
  ylab("Proportion (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  plot_theme("dark")
```

:::

```{r}
cat("title=", "OA rate by institution", paste0("(", prev_year, ")"))
  
pubs_all |> 
  filter(country_code == "AU", 
         publication_year == max(publication_year)) |>
  group_by(institution, is_oa) |> 
  tally() |> 
  pivot_wider(names_from = is_oa, values_from = n) |> 
  mutate(pc_open = `TRUE` / (`TRUE` + `FALSE`)) |> 
  arrange(desc(pc_open)) |> 
  select(-`FALSE`, -`TRUE`) |>
  ungroup() |> 
  gt() |> 
  fmt_percent(decimals = 0) |> 
  tab_options(column_labels.hidden = TRUE,
              table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::

::::::

### Comparison

:::::: {.rows}

```{r}
#| title: NZ vs AU OA rate over time

annual_compare <- 
  full_join(get_annual_open("NZ"), 
            get_annual_open("AU")) |> 
  mutate(
    country_code = case_when(row_number() %% 2 == 0 ~ "AU", .default = "NZ")
    )

ggplot(annual_compare, aes(x = publication_year, y = freq)) +
  geom_line(data = annual_compare |> filter(country_code == "NZ"), 
            size = 1, aes(colour = "NZ")) +
  geom_line(data = annual_compare |> filter(country_code == "AU"), 
            size = 1, aes(colour = "AU")) +
  theme_classic() +
  scale_color_manual(name = "Country", values = c("NZ" = "#f3f2f2", "AU" = "gold"),
                     guide = guide_legend(reverse = TRUE)) +
  xlab("") +
  ylab("Open Access (%)") +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  scale_x_continuous(n.breaks = prev_year-min(annual_compare$publication_year)) +
  plot_theme("dark")
```

::::::

# Institutions

## Column {width=50%}

```{r}
#| title: Institutional Trend (NZ)

pubs_all |> 
  filter(country_code == "NZ") |> 
  group_by(publication_year, institution, is_oa) |>
  tally() |>
  pivot_wider(names_from = is_oa, values_from = n) |>
  mutate(pc_open = `TRUE` / (`TRUE` + `FALSE`)) |> 
  plot_trend() +
  scale_colour_manual(values = inst_lines_nz) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Row {.tabset}

```{r}
inst_nz <- data.frame(institution = unique(pubs_all$institution),
                   abbrev = unique(pubs_all$abbrev))

plot_nz_list <- 
  pubs_all |>
  filter(country_code == "NZ") |> 
  group_by(publication_year, institution, oa_status) |>
  tally() |>
  mutate(
    pc = n / sum(n)
  ) |> 
  left_join(.x, y = inst_nz, by = join_by(institution)) |> 
  group_by(institution) |>
  group_split() |>
  map(~ plot_inst(.x, "dark"))
```

```{r}
cat("title=", plot_nz_list[[1]][["data"]][["abbrev"]][1])
plot_nz_list[[1]]
```

```{r}
cat("title=", plot_nz_list[[2]][["data"]][["abbrev"]][1])
plot_nz_list[[2]]
```

```{r}
cat("title=", plot_nz_list[[3]][["data"]][["abbrev"]][1])
plot_nz_list[[3]]
```

```{r}
cat("title=", plot_nz_list[[4]][["data"]][["abbrev"]][1])
plot_nz_list[[4]]
```

```{r}
cat("title=", plot_nz_list[[5]][["data"]][["abbrev"]][1])
plot_nz_list[[5]]
```

```{r}
cat("title=", plot_nz_list[[6]][["data"]][["abbrev"]][1])
plot_nz_list[[6]]
```

```{r}
cat("title=", plot_nz_list[[7]][["data"]][["abbrev"]][1])
plot_nz_list[[7]]
```

```{r}
cat("title=", plot_nz_list[[8]][["data"]][["abbrev"]][1])
plot_nz_list[[8]]
```

## Column {width=50%}

```{r}
inst_au <- data.frame(institution = unique(pubs_all$institution),
                   abbrev = unique(pubs_all$abbrev))

plot_au_list <- 
  pubs_all |>
  filter(country_code == "AU") |> 
  group_by(publication_year, institution, oa_status) |>
  tally() |>
  mutate(
    pc = n / sum(n)
  ) |> 
  left_join(.x, y = inst_au, by = join_by(institution)) |> 
  group_by(institution) |>
  group_split() |>
  map(~ plot_inst(.x, "dark"))
```

```{r}
#| title: Institutional Trend (AU)

pubs_all |> 
  filter(country_code == "AU") |> 
  group_by(publication_year, institution, is_oa) |>
  tally() |>
  pivot_wider(names_from = is_oa, values_from = n) |>
  mutate(pc_open = `TRUE` / (`TRUE` + `FALSE`)) |> 
  plot_trend() +
  scale_colour_manual(values = inst_lines_au) +
  scale_y_continuous(labels = scales::label_percent(),
                     limits = c(0, 0.8),
                     breaks = c(0, 0.2, 0.4, 0.6, 0.8)) +
  plot_theme("dark") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```

### Row {.tabset}

```{r}
cat("title=", plot_au_list[[1]][["data"]][["abbrev"]][1])
plot_au_list[[1]]
```

```{r}
cat("title=", plot_au_list[[2]][["data"]][["abbrev"]][1])
plot_au_list[[2]]
```

```{r}
cat("title=", plot_au_list[[3]][["data"]][["abbrev"]][1])
plot_au_list[[3]]
```

```{r}
cat("title=", plot_au_list[[4]][["data"]][["abbrev"]][1])
plot_au_list[[4]]
```

```{r}
cat("title=", plot_au_list[[5]][["data"]][["abbrev"]][1])
plot_au_list[[5]]
```

```{r}
cat("title=", plot_au_list[[6]][["data"]][["abbrev"]][1])
plot_au_list[[6]]
```

```{r}
cat("title=", plot_au_list[[7]][["data"]][["abbrev"]][1])
plot_au_list[[7]]
```

```{r}
cat("title=", plot_au_list[[8]][["data"]][["abbrev"]][1])
plot_au_list[[8]]
```

# Field

## Row {.tabset}

### NZ

```{r}
#| title: "OA Rate By Field (NZ)"

pubs_all |>
  filter(country_code == "NZ") |> 
  drop_na(field) |> 
  group_by(field, is_oa) |> 
  tally() |> 
  mutate(
    pc = n / sum(n)
    ) |>
  filter(is_oa == TRUE) |>
  ggplot(aes(x = reorder(field, pc), y = pc)) +
  geom_bar(position = "dodge", stat = "identity", fill = "#4DAF4A") +
  geom_text(aes(label = n, y = pc - 0.02),
            vjust = 0.5,
            colour = "white",
            size = 3) +
  coord_flip() +
  plot_theme("dark") +
  ylab("") +
  xlab("") +
  labs(caption = "Numbers on bars are counts of open works.") +
  scale_y_continuous(labels = label_percent(),
                     limits = c(0, 0.8)) +
  theme(text = element_text(size = 10))
```

### AU

```{r}
#| title: "OA Rate By Field (AU)"

pubs_all |>
  filter(country_code == "AU") |> 
  drop_na(field) |> 
  group_by(field, is_oa) |> 
  tally() |> 
  mutate(
    pc = n / sum(n)
    ) |>
  filter(is_oa == TRUE) |>
  ggplot(aes(x = reorder(field, pc), y = pc)) +
  geom_bar(position = "dodge", stat = "identity", fill = "#4DAF4A") +
  geom_text(aes(label = n, y = pc - 0.02),
            vjust = 0.5,
            colour = "white",
            size = 3) +
  coord_flip() +
  plot_theme("dark") +
  ylab("") +
  xlab("") +
  labs(caption = "Numbers on bars are counts of open works.") +
  scale_y_continuous(labels = label_percent(),
                     limits = c(0, 0.9)) +
  theme(text = element_text(size = 10))
```

# Citations

## Row {.tabset}

### NZ

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Publications"

make_valuebox("file-text", "info", 
              format(nrow(pubs_all |> filter(country_code == "NZ")), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Citations"

make_valuebox("blockquote-left", "info", 
              format(
                pubs_all |> 
                  filter(country_code == "NZ") |> 
                  summarise(citations = sum(cited_by_count)) |> 
                  pull(citations), 
                big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant_nz <-
pubs_all |> 
  filter(country_code == "NZ") |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

make_valuebox("bar-chart", "success", scales::percent(1 - cit_advant_nz$difference))
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", 
              format(file.info("data/pubs-all.rds")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: "Open vs Closed"

pubs_all |>
  filter(country_code == "NZ") |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

```{r}
#| title: "Open Access Status"

pubs_all |>
  filter(country_code == "NZ") |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

:::

::: {.rows}

```{r}
#| title: "Citation Summary"

pubs_all |> 
  filter(country_code == "NZ") |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread"

pubs_all |>
  filter(country_code == "NZ") |> 
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks = c(0,1,2,5,10,100,1000), trans = "log1p", expand = c(0,0)) +
  plot_theme("dark") +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```

:::

::::

::::::

### AU

:::::: {.rows}

::::: {.columns}

```{r}
#| content: valuebox
#| title: "Publications"

make_valuebox("file-text", "info", 
              format(nrow(pubs_all |> filter(country_code == "AU")), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Citations"

make_valuebox("blockquote-left", "info", 
              format(
                pubs_all |> 
                  filter(country_code == "AU") |> 
                  summarise(citations = sum(cited_by_count)) |> 
                  pull(citations), 
                big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "OA Citation Advantage"

cit_advant_nz <-
pubs_all |> 
  filter(country_code == "AU") |> 
  group_by(is_oa) |> 
  summarise(
    med_citations = median(cited_by_count)
  ) |> 
  pivot_wider(names_from = is_oa, values_from = med_citations) |> 
  mutate(
    difference = `FALSE` / `TRUE`,
  )

make_valuebox("bar-chart", "success", scales::percent(1 - cit_advant_nz$difference))
```

```{r}
#| content: valuebox
#| title: "Last Updated"

make_valuebox("clock-history", "light", format(file.info("data/pubs-all.rds")$mtime, format = "%b %Y"))
```

:::::

:::: {.columns}

::: {.rows}

```{r}
#| title: "Open vs Closed"

pubs_all |>
  filter(country_code == "AU") |> 
  group_by(publication_year, is_oa) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  mutate(
    access = case_when(is_oa == "TRUE" ~ "Open",
                       .default = "Closed")
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = access)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "", values = c("Open" = "#4DAF4A", "Closed" = "#E41A1C"),
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

```{r}
#| title: "Open Access Status"

pubs_all |>
  filter(country_code == "AU") |> 
  group_by(publication_year, oa_status) |>
  summarise(
    median = median(cited_by_count)
  ) |> 
  ggplot(aes(x = publication_year, y = median, fill = oa_status)) +
  geom_bar(position="dodge", stat="identity") +
  coord_flip() +
  plot_theme("dark") +
  scale_fill_manual(name = "Access Type", values = plot_colours, 
                    guide = guide_legend(reverse = TRUE)) +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  xlab("") +
  ylab("Median Citations")
```

:::

::: {.rows}

```{r}
#| title: "Citation Summary"

pubs_all |> 
  filter(country_code == "AU") |> 
  group_by(publication_year) |> 
  summarise(
    Works = format(n(), big.mark = ","),
    Citations = format(sum(cited_by_count), big.mark = ","),
    Min = min(cited_by_count),
    Max = format(max(cited_by_count), big.mark = ","),
    Median = median(cited_by_count),
    Average = round(mean(cited_by_count), digits = 1)
  ) |>
  gt() |> 
  cols_label(publication_year = "Year") |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

```{r}
#| title: "Citation Spread"

pubs_all |>
  filter(country_code == "AU") |> 
  ggplot(aes(x = cited_by_count)) +
  stat_density(aes(y = after_stat(count)), color = "black", fill = "#f3f2f2") +
  scale_x_continuous(breaks = c(0,1,2,5,10,100,1000), trans = "log1p", expand = c(0,0)) +
  plot_theme("dark") +
  xlab("Citations (log1p)") +
  ylab("Works") +
  scale_y_continuous(labels = comma)
```

:::

::::

::::::

# Publishers

## Row {.tabset}

### NZ

:::: {.columns}

::: {.rows}

```{r clean-publishers}
pubs_all <-
  pubs_all |> 
  mutate(
    publisher_clean = case_when(
      str_detect(publisher, "Elsevier|RELX|Cell Press") ~ "Elsevier",
      str_detect(publisher, "Wiley") ~ "Wiley",
      str_detect(publisher, "Multidisciplinary Digital Publishing Institute") ~ "MDPI",
      str_detect(publisher, "Taylor & Francis|Routledge|Informa") ~ "Taylor & Francis",
      str_detect(publisher, "Springer|Nature Portfolio") ~ "Springer",
      str_detect(publisher, "Frontiers") ~ "Frontiers Media",
      str_detect(publisher, "IEEE") ~ "IEEE",
      str_detect(publisher, "Institute of Electrical and Electronics Engineers") ~ "IEEE",
      str_detect(publisher, "De Gruyter|Brill") ~ "De Gruyter Brill",
      str_detect(publisher, "Q15088586") ~ "Magnolia Press",
      str_detect(publisher, "Oxford") ~ "Oxford University Press",
      str_detect(publisher, "Wolters|Lippincott") ~ "Wolters Kluwer",
      str_detect(publisher, "Massachusetts Institute of Technology|MIT Press") ~ "The MIT Press",
      .default = publisher)
    )
```

```{r}
#| title: Top 5 Share Over Time

top5 <- 
  pubs_all |> 
  filter(country_code == "NZ", type == "article") |> 
  drop_na(publisher_clean) |> 
  group_by(publisher_clean) |> 
  tally() |> 
  arrange(desc(n)) |> 
  slice_head(n = 5) |> 
  pull(publisher_clean)

pubs_all |> 
  filter(country_code == "NZ", type == "article") |> 
  mutate(
    publisher_clean2 = case_when(
      publisher_clean %in% top5 ~ publisher_clean,
      .default = "Other"
    )
  ) |> 
  group_by(publication_year, publisher_clean2) |> 
  tally() |> 
  mutate(
    pc = n / sum(n),
  ) |> 
  select(-n) |>
  ggplot(aes(x = publication_year, y = pc, fill = publisher_clean2)) +
  geom_area() +
  xlab("") +
  ylab("Article Share (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_nz) +
  plot_theme("dark")
```

```{r}
#| title: Top 5 By Field

pubs_all |>
  filter(country_code == "NZ", type == "article") |>
  drop_na(field) |>
  filter(publisher_clean %in% top5) |>
  group_by(publisher_clean, field) |>
  summarise(
    n = n()
  ) |>
  mutate(
    pc = n/ sum(n)
  ) |>
  ggplot(aes(x = field, y = pc, fill = publisher_clean)) +
  geom_bar(position = "fill", stat = "identity") +
  coord_flip() +
  xlab("") +
  ylab("") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_nz[1:5]) +
  plot_theme("dark")
```

:::

```{r}
#| title: Top 20 by Article Share

publisher_rates <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "NZ", type == "article") |> 
  group_by(publisher_clean, is_oa) |> 
  tally() |> 
  mutate(
    percent_open = n / sum(n)
  ) |> 
  filter(is_oa == TRUE) |> 
  arrange(desc(n))

publisher_citations <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "NZ", type == "article") |> 
  group_by(publisher_clean) |> 
  summarise(
    citations = sum(cited_by_count),
    articles = n(),
    median_citations = median(cited_by_count)
  ) |> 
  filter(articles > 20) |> 
  arrange(desc(median_citations))

nz_articles <- nrow(filter(pubs_all, country_code == "NZ", type == "article"))

left_join(publisher_rates, publisher_citations) |> 
  rename(Publisher = publisher_clean,
         Articles = articles,
         "Percent OA" = percent_open,
         "Median Citations" = median_citations) |> 
  select(-is_oa, -n, -citations) |> 
  relocate(c(Articles, "Percent OA", "Median Citations"), .after = Publisher) |>
  arrange(desc(Articles)) |>
  mutate(
    "Article Share" = Articles / nz_articles
  ) |> 
  ungroup() |> 
  slice_head(n = 20) |> 
  mutate(
    Rank = row_number()
  ) |>
  relocate(Rank) |> 
  relocate("Article Share", .after = Articles) |> 
  gt() |> 
  fmt_number(columns = c(Articles, "Median Citations"), 
             sep_mark = ",", decimals = 0) |> 
  fmt_percent(columns = c("Percent OA", "Article Share"), decimals = 0) |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::

### AU

:::: {.columns}

::: {.rows}

```{r}
#| title: Top 5 Share Over Time

top5 <- 
  pubs_all |> 
  filter(country_code == "AU", type == "article") |> 
  drop_na(publisher_clean) |> 
  group_by(publisher_clean) |> 
  tally() |> 
  arrange(desc(n)) |> 
  slice_head(n = 5) |> 
  pull(publisher_clean)

pubs_all |> 
  filter(country_code == "AU", type == "article") |> 
  mutate(
    publisher_clean2 = case_when(
      publisher_clean %in% top5 ~ publisher_clean,
      .default = "Other"
    )
  ) |> 
  group_by(publication_year, publisher_clean2) |> 
  tally() |> 
  mutate(
    pc = n / sum(n),
  ) |> 
  select(-n) |>
  ggplot(aes(x = publication_year, y = pc, fill = publisher_clean2)) +
  geom_area() +
  xlab("") +
  ylab("Article Share (%)") +
  scale_x_continuous(n.breaks = prev_year-min(pubs_all$publication_year)) +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_au) +
  plot_theme("dark")
```

```{r}
#| title: Top 5 By Field

pubs_all |>
  filter(country_code == "AU", type == "article") |>
  drop_na(field) |>
  filter(publisher_clean %in% top5) |>
  group_by(publisher_clean, field) |>
  summarise(
    n = n()
  ) |>
  mutate(
    pc = n/ sum(n)
  ) |>
  ggplot(aes(x = field, y = pc, fill = publisher_clean)) +
  geom_bar(position = "fill", stat = "identity") +
  coord_flip() +
  xlab("") +
  ylab("") +
  scale_y_continuous(labels = scales::label_percent()) +
  scale_fill_manual(values = publisher_colours_au[1:5]) +
  plot_theme("dark")
```

:::

```{r}
#| title: Top 20 by Article Share

publisher_rates <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "AU", type == "article") |> 
  group_by(publisher_clean, is_oa) |> 
  tally() |> 
  mutate(
    percent_open = n / sum(n)
  ) |> 
  filter(is_oa == TRUE) |> 
  arrange(desc(n))

publisher_citations <-
pubs_all |> 
  drop_na(publisher_clean) |> 
  filter(country_code == "AU", type == "article") |> 
  group_by(publisher_clean) |> 
  summarise(
    citations = sum(cited_by_count),
    articles = n(),
    median_citations = median(cited_by_count)
  ) |> 
  filter(articles > 20) |> 
  arrange(desc(median_citations))

nz_articles <- nrow(filter(pubs_all, country_code == "AU", type == "article"))

left_join(publisher_rates, publisher_citations) |> 
  rename(Publisher = publisher_clean,
         Articles = articles,
         "Percent OA" = percent_open,
         "Median Citations" = median_citations) |> 
  select(-is_oa, -n, -citations) |> 
  relocate(c(Articles, "Percent OA", "Median Citations"), .after = Publisher) |>
  arrange(desc(Articles)) |>
  mutate(
    "Article Share" = Articles / nz_articles
  ) |> 
  ungroup() |> 
  slice_head(n = 20) |> 
  mutate(
    Rank = row_number()
  ) |>
  relocate(Rank) |> 
  relocate("Article Share", .after = Articles) |> 
  gt() |> 
  fmt_number(columns = c(Articles, "Median Citations"), 
             sep_mark = ",", decimals = 0) |> 
  fmt_percent(columns = c("Percent OA", "Article Share"), decimals = 0) |> 
  tab_options(table.background.color = "#2d2d2d",
              table.font.color.light = "white")
```

::::