---
title: "OA Dashboard"
author: "Tom Saunders"
output:
  html_document:
    css: "style.css"
knit: (function(inputFile, encoding) { rmarkdown::render(inputFile, encoding = encoding, output_file = file.path(dirname(inputFile), 'index.html')) })
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo=FALSE)
knitr::opts_chunk$set(warning=FALSE)
knitr::opts_chunk$set(message=FALSE)
knitr::opts_chunk$set(fig.align="center")
library(tidyverse)
```

This dashboard displays access type data for the scholarly publication output of all 8 Aotearoa New Zealand universities to track progress towards the Universities New Zealand [target](https://www.universitiesnz.ac.nz/sites/default/files/uni-nz/documents/Open%20Access%20Statement.pdf) of 70% open access output by 2025. It also includes data from the Group of 8 Australian universities to examine how our two countries compare. 

Data is extracted from OpenAlex via the API to examine access trends since 2010 using two queries:

* [This query](https://github.com/tesaunders/oa-nz/blob/main/query-overall.R) extracts data relating to the overall access trend for both sets of universities.
* [This query](https://github.com/tesaunders/oa-nz/blob/main/query-institution.R) extracts data relating to individual universities. 

All code, plots, and summarised data can be found [here](https://github.com/tesaunders/oa-nz).

## Overall Data

```{r, fig.dim = c(8, 6)}
# Get previous year

prev_year <- as.numeric(format(Sys.Date(), "%Y"))-1

# Read in data generated by both queries

data_overall_nz <- read_csv("data/overall_nz.csv") |> 
  mutate(country = "nz")

data_overall_au <- read_csv("data/overall_au.csv") |> 
  mutate(country = "au")

data_institutions_nz <- read_csv("data/institutions_nz.csv") |> 
  mutate(country = "nz")

data_institutions_au <- read_csv("data/institutions_au.csv") |> 
  mutate(country = "au")

data_overall <- rbind(data_overall_nz, data_overall_au)
data_institutions <- rbind(data_institutions_nz, data_institutions_au)


# Summarise and plot overall trend data

overall_trend <- data_overall |>
  mutate(
    open = case_when(oa_type != "closed" ~ "open", .default = oa_type),
  ) |> 
  group_by(year, open, country) |> 
  summarise(
    pc = sum(pc),
  ) |> 
  filter(open == "open")

overall_trend_title <- paste0("Open Access Publication Output, 2010-", prev_year)

ggplot(overall_trend, aes(x = year, y = pc)) +
  geom_line(data = overall_trend |> filter(country == "nz"), size = 1, aes(colour = "NZ")) +
  geom_line(data = overall_trend |> filter(country == "au"), size = 1, aes(colour = "AU")) +
  theme_classic() +
  scale_color_manual(name = "Country",
                     values = c("NZ" = "black",
                                "AU" = "gold")) +
  xlab("") +
  ylab("%") +
  ylim(0, 80) +
  scale_x_continuous(n.breaks = prev_year-2010) +
  ggtitle(overall_trend_title)

ggsave("fig/1-overall-trend.png")

# Plot break down of access type

data_overall$oa_type <- factor(data_overall$oa_type, 
                             levels = c("bronze","green", "hybrid", "gold", "closed"))

overall_types_title_nz <- (paste0("Access Status for New Zealand Research Publications, 2010-", prev_year))
overall_types_title_au <- (paste0("Access Status for Australian Research Publications, 2010-", prev_year))

ggplot(data_overall, aes(x = year, y = pc, fill = oa_type)) +
  geom_bar(data = data_overall |> filter(country == "nz"),
           position = "dodge", stat = "identity") +
  scale_fill_manual(name = "Access Type", 
                     values = c("bronze" = "#A65628",
                                "hybrid" = "#377EB8",
                                "gold" = "#FFFF33",
                                "green" = "#4DAF4A",
                                "closed" = "#E41A1C")) +
  theme_classic() +
  xlab("") +
  ylab("%") +
  scale_x_continuous(n.breaks = prev_year-2010) +
  ggtitle(overall_types_title_nz)

ggsave("fig/2-overall-types-nz.png")

ggplot(data_overall, aes(x = year, y = pc, fill = oa_type)) +
  geom_bar(data = data_overall |> filter(country == "au"),
           position = "dodge", stat = "identity") +
  scale_fill_manual(name = "Access Type", 
                     values = c("bronze" = "#A65628",
                                "hybrid" = "#377EB8",
                                "gold" = "#FFFF33",
                                "green" = "#4DAF4A",
                                "closed" = "#E41A1C")) +
  theme_classic() +
  xlab("") +
  ylab("%") +
  scale_x_continuous(n.breaks = prev_year-2010) +
  ggtitle(overall_types_title_au)

ggsave("fig/2-overall-types-au.png")
```

## Comparison of Institutions {.tabset}

### Open status over time

```{r, fig.dim = c(8, 60)}
data_institutions_sum <- data_institutions |> 
  mutate(
    open = case_when(oa_type != "closed" ~ "open", .default = oa_type),
  ) |> 
  group_by(year, institution, open) |> 
  summarise(
    pc = sum(pc),
  )

institution_title <- paste0("Open Access Publication Output for ", sort(unique(data_institutions$institution)), ", 2010-", prev_year)

plot_function_open <- function(data_institutions_sum) {
  p <- ggplot(data_institutions_sum, aes(x = year, y = pc, colour = open))
  p <- p + geom_line(size = 1) +
    scale_colour_manual(name = "Access Type", 
                        values = c("open" = "#4DAF4A",
                                   "closed" = "#E41A1C")) +
    theme_classic() +
    xlab("") +
    ylab("%") +
    ylim(0, 80) +
    scale_x_continuous(n.breaks = prev_year-2010)
  return(p)
}

plot_list <- data_institutions_sum |>
  group_by(institution) |> 
  group_split(institution)

plot_list <- plot_list[order(sapply(plot_list, function(x) sort(unique(x$institution))))]

plot_list <- plot_list |> 
map(~ plot_function_open(.x))

plot_list <- lapply(seq_along(plot_list), function(i) { 
  plot_list[[i]] + ggtitle(institution_title[i])
})

wrapped_plot <- patchwork::wrap_plots(plot_list, ncol = 1)

nb_plots <- length(plot_list)
baseheight <- grDevices::dev.size(units = "px")[1]
ggsave("fig/3-institutions-all.png",
       wrapped_plot,
       height = nb_plots * baseheight,
       units = "px",
       limitsize = FALSE)

wrapped_plot
```

### Access type over time

```{r, fig.dim = c(8, 60)}
institution_type_title <- paste0("Access Type Trend for ", sort(unique(data_institutions$institution)), ", 2010-", prev_year)

plot_function_type <- function(data_institutions) {
  p <- ggplot(data_institutions, aes(x = year, y = pc, colour = oa_type))
  p <- p + geom_line(size = 1) +
    scale_colour_manual(name = "Access Type", 
                       values = c("bronze" = "#A65628",
                                  "hybrid" = "#377EB8",
                                  "gold" = "#FFFF33",
                                  "green" = "#4DAF4A",
                                  "closed" = "#E41A1C")) +
    theme_classic() +
    xlab("") +
    ylab("%") +
    ylim(0, 80) +
    scale_x_continuous(n.breaks = prev_year-2010)
  return(p)
}

plot_list_type <- data_institutions |>
  group_by(institution) |>
  group_split()

plot_list_type <- plot_list_type[order(sapply(plot_list_type, 
                                              function(x) sort(unique(x$institution))))]
plot_list_type <- plot_list_type |> 
map(~ plot_function_type(.x))

plot_list_type <- lapply(seq_along(plot_list_type), function(i) { 
  plot_list_type[[i]] + ggtitle(institution_type_title[i])
})

wrapped_plot_type <- patchwork::wrap_plots(plot_list_type, ncol = 1)

nb_plots_type <- length(plot_list_type)
ggsave("fig/4-institutions-type.png",
       wrapped_plot_type,
       height = nb_plots_type * baseheight,
       units = "px",
       limitsize = FALSE)

wrapped_plot_type
```

### Access type for most recent year

```{r, fig.dim = c(8, 6)}
institutions_access <- data_institutions |> 
  filter(year == prev_year) 

order_closed <- institutions_access |> 
  filter(oa_type == "closed") |> 
  arrange(desc(pc)) |> 
  select(institution)

institutions_access$institution <- factor(institutions_access$institution, 
                                          levels = pull(order_closed, institution))

institutions_access$oa_type <- factor(institutions_access$oa_type, 
                             levels = c("bronze","green", "hybrid", "gold", "closed"))

institutions_access_title <- (paste0("Access Type by Institution, ", prev_year))

ggplot(institutions_access, aes(x = institution, y = pc, fill = oa_type)) +
  geom_bar(position = "dodge", stat = "identity") +
  scale_fill_manual(name = "Access Type", 
                     values = c("bronze" = "#A65628",
                                "hybrid" = "#377EB8",
                                "gold" = "#FFFF33",
                                "green" = "#4DAF4A",
                                "closed" = "#E41A1C")) +
  coord_flip() +
  theme_classic() +
  xlab("") +
  ylab("%") +
  ggtitle(institutions_access_title)

ggsave("fig/5-institutions-type-now.png")

```

##  {.unnumbered}